---
title: "OmniCellAgent case study: cell-level DE and panels A-D"
output: html_document
date: "`r Sys.Date()`"
params:
  session_dir: "/storage3/fs1/fuhai.li/Active/di.huang/Research/LLM/OmniCellAgent/webapp/sessions/alzheimer_test_910"
  ref_csv: "foranalysis_combined_normal_df_Alzheimer's Disease.csv"
  alt_csv: "foranalysis_combined_disease_df_Alzheimer's Disease.csv"
  ref_label: "Healthy"
  alt_label: "Diseased"
  out_dir: null
  out_prefix: "DE_results"
  input_scale: "linear_cp10k"
  fdr_thresh: 0.05
  fc_thresh: 0.25
  permutations: 999
  n_corrgram: 20
  n_pca: 100
  which_pc: "PC1"
  composite: true
  do_pseudobulk: false
  labels_csv: "labels.csv"
---

```{r setup, include=FALSE, dev="pdf"}
# dev="pdf" on THIS chunk is required, not cosmetic: knitr opens a device before
# evaluating a chunk, so the opts_chunk$set(dev=) below cannot apply to the
# chunk it lives in. Without it, knitr falls back to svg here and dies with
# "Cairo-based devices are not available for this platform". pdf() is the only
# device compiled into this R build.
#
# The r/4.5.2 module on this cluster is built WITHOUT png, jpeg, cairo or X11:
#   capabilities()[c("png","jpeg","cairo","X11")]  ->  FALSE FALSE FALSE FALSE
# so base png() raises "X11 is not available" and knitr's default device fails
# on every chunk, not just the plotting ones. ragg supplies PNG independently of
# R's compiled-in devices, and pdf() works natively. Do not replace this with
# png() without re-checking capabilities() on the machine you are running on.
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE, dev = "ragg_png")

suppressPackageStartupMessages({
  library(ragg)
  library(ggplotify)
  library(limma)
  library(ggplot2)
  library(ggrepel)
  library(patchwork)
  library(RColorBrewer)
  library(data.table)
  library(corrgram)
  library(vegan)
  library(multcompView)
  library(ggpubr)
  library(pheatmap)
})

out_dir <- if (is.null(params$out_dir)) params$session_dir else params$out_dir
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)
out_stem <- file.path(out_dir, params$out_prefix)

fdr_thresh <- params$fdr_thresh
fc_thresh  <- params$fc_thresh

if (!capabilities("png")) {
  message("[device] R has no built-in png device; using ragg for all raster output.")
}
```

## What this document does

Reads the two `foranalysis_combined_*` matrices a session emits, runs limma on
the **cell** axis, and writes panels A-D as separate image files. The four panels
are assembled into a single figure outside R.

Read the caveats in "Statistical scope" at the bottom before using any of this
for an inferential claim.

## 1. Load and align

Each CSV is genes x cells: column 1 is `Name` (HGNC symbol), every other column
is one **cell** (`ns_sample_i` / `ds_sample_i`). Despite the name, a column is
not a biological sample.

```{r load-data}
read_expr <- function(path) {
  stopifnot(file.exists(path))
  dt <- fread(path)
  if (!is.numeric(dt[[1]])) {
    rn <- dt[[1]]
    dt[, 1 := NULL]
    mat <- as.matrix(dt)
    rownames(mat) <- rn
  } else {
    mat <- as.matrix(dt)
    if (is.null(rownames(mat))) rownames(mat) <- paste0("gene_", seq_len(nrow(mat)))
  }
  storage.mode(mat) <- "numeric"
  mat
}

mat_ref <- read_expr(file.path(params$session_dir, params$ref_csv))
mat_alt <- read_expr(file.path(params$session_dir, params$alt_csv))

message(sprintf("%s: %d genes x %d cells", params$ref_label, nrow(mat_ref), ncol(mat_ref)))
message(sprintf("%s: %d genes x %d cells", params$alt_label, nrow(mat_alt), ncol(mat_alt)))
```

## 2. Verify the input scale, then transform

The `q99 > 100` heuristic the earlier version used is not a scale detector: a
maximum around 5 is equally consistent with log-scale data and with low-valued
linear normalised expression.

These matrices come out of `normalize_cp10k()` in
`omic_fetch_analysis_workflow.py`, which rescales every **cell** to sum to
10,000 and does **not** apply `log1p`. That gives a structural signature rather
than a magnitude guess: each column must sum to 1e4. `params$input_scale`
declares the expected scale; this chunk asserts it and fails loudly on a
mismatch instead of silently double-logging.

```{r scale-check}
describe_scale <- function(m, nm) {
  cs <- colSums(m, na.rm = TRUE)
  q  <- quantile(m, c(0.5, 0.9, 0.99), na.rm = TRUE)
  message(sprintf(
    "[scale] %-10s median colSum = %.1f | q50 = %.4g  q90 = %.4g  q99 = %.4g | max = %.4g",
    nm, median(cs), q[[1]], q[[2]], q[[3]], max(m, na.rm = TRUE)))
  median(cs)
}

med_ref <- describe_scale(mat_ref, params$ref_label)
med_alt <- describe_scale(mat_alt, params$alt_label)

if (identical(params$input_scale, "linear_cp10k")) {
  tol <- 0.01
  ok <- abs(med_ref - 1e4) / 1e4 < tol && abs(med_alt - 1e4) / 1e4 < tol
  if (!ok) {
    stop(sprintf(paste0(
      "input_scale is declared 'linear_cp10k' but median column sums are ",
      "%.1f / %.1f, not 10000. Either the matrices were already transformed, ",
      "or the upstream normalisation changed. Fix params$input_scale rather ",
      "than letting log2() run on the wrong scale."), med_ref, med_alt))
  }
  message("[scale] CP10K confirmed (columns sum to 1e4). Applying log2(x + 1).")
  transform_applied <- TRUE
} else if (identical(params$input_scale, "log2_already")) {
  message("[scale] Caller declares data is already log2. No transform applied.")
  transform_applied <- FALSE
} else {
  stop("params$input_scale must be 'linear_cp10k' or 'log2_already'.")
}
```

`limma` assumes a log scale. Without the transform, `logFC` is a difference of
linear CP10K means and cannot be called a log2 fold change.

```{r combine}
common_genes <- intersect(rownames(mat_ref), rownames(mat_alt))
if (length(common_genes) < 10) stop("Too few overlapping genes between files.")
mat_ref <- mat_ref[common_genes, , drop = FALSE]
mat_alt <- mat_alt[common_genes, , drop = FALSE]

expr <- cbind(mat_ref, mat_alt)
group <- factor(
  c(rep(params$ref_label, ncol(mat_ref)), rep(params$alt_label, ncol(mat_alt))),
  levels = c(params$ref_label, params$alt_label)
)
colnames(expr) <- make.unique(colnames(expr), sep = "_")

if (transform_applied) expr <- log2(expr + 1)

# Drop mostly-missing and zero-variance genes
expr <- expr[rowSums(is.finite(expr)) >= ncol(expr) * 0.8, , drop = FALSE]
v <- apply(expr, 1, stats::var, na.rm = TRUE)
expr <- expr[v > 0, , drop = FALSE]

message(sprintf("Analysis matrix: %d genes x %d cells (%s n=%d, %s n=%d)",
                nrow(expr), ncol(expr),
                params$ref_label, sum(group == params$ref_label),
                params$alt_label, sum(group == params$alt_label)))
```

## 3. Differential expression

Means parameterisation (`~ 0 + group`) with an explicit contrast. Note that
`~ 0 + group` **must** be paired with `makeContrasts()` / `contrasts.fit()`:
its coefficients are the two group means, so testing `coef = "groupDiseased"`
directly would test "mean expression differs from zero", which is true for
essentially every expressed gene. The contrast below is the alt-minus-ref
difference and is what `logFC` reports.

```{r de}
design <- model.matrix(~ 0 + group)
colnames(design) <- make.names(levels(group))

contrast_str <- paste(make.names(params$alt_label), "-", make.names(params$ref_label))
cm <- makeContrasts(contrasts = contrast_str, levels = design)

fit  <- lmFit(expr, design)
fit2 <- contrasts.fit(fit, cm)
fit2 <- eBayes(fit2, robust = TRUE, trend = TRUE)

res <- topTable(fit2, coef = 1, number = Inf, sort.by = "P",
                adjust.method = "BH")
res$Gene <- rownames(res)
res <- res[, c("Gene", "logFC", "AveExpr", "t", "P.Value", "adj.P.Val", "B")]

out_csv <- paste0(out_stem, "_table.csv")
fwrite(res, out_csv)
message("Contrast: ", contrast_str)
message("Results written to: ", out_csv)

sig_count <- sum(res$adj.P.Val < fdr_thresh, na.rm = TRUE)
message("Genes with FDR < ", fdr_thresh, ": ", sig_count)
if (sig_count > 0) {
  message("Up in ", params$alt_label, ": ",
          sum(res$adj.P.Val < fdr_thresh & res$logFC > 0, na.rm = TRUE),
          " | Down: ",
          sum(res$adj.P.Val < fdr_thresh & res$logFC < 0, na.rm = TRUE))
}
```

## 4. PCA and PERMANOVA (calculations only)

Split out so no plotting chunk silently creates data another plotting chunk
needs. Panels C and D both consume `pcoadata` produced here.

```{r pca-permanova}
top_genes_pca <- head(res$Gene[order(res$adj.P.Val)], n = min(params$n_pca, nrow(res)))

pca <- prcomp(t(expr[top_genes_pca, , drop = FALSE]), scale. = TRUE)
pve <- (pca$sdev^2) / sum(pca$sdev^2)

pcoadata <- data.frame(
  sample  = colnames(expr),
  PC1     = pca$x[, 1],
  PC2     = pca$x[, 2],
  Subtype = group
)

pal2 <- setNames(colorRampPalette(brewer.pal(12, "Paired"))(2),
                 c(params$ref_label, params$alt_label))

# Bray-Curtis needs non-negative input; log2(CP10K + 1) satisfies that.
dist_mat   <- vegdist(t(expr[top_genes_pca, , drop = FALSE]), method = "bray")
otu.adonis <- adonis2(dist_mat ~ Subtype, data = pcoadata,
                      permutations = params$permutations)

# A permutation test cannot resolve p below 1/(permutations + 1). Report the
# floor as an inequality rather than as a measured probability.
perm_floor <- 1 / (params$permutations + 1)
p_obs <- otu.adonis$`Pr(>F)`[1]
perm_p_label <- if (!is.na(p_obs) && p_obs <= perm_floor + 1e-12) {
  paste0("p <= ", format(perm_floor, scientific = FALSE))
} else {
  paste0("p = ", signif(p_obs, 3))
}
message("PERMANOVA (", params$permutations, " permutations): R2 = ",
        round(otu.adonis$R2[1], 5), ", ", perm_p_label)
```

## 5. Panel A -- volcano

The y-axis and point significance both use `adj.P.Val`, which `topTable()`
calculates by applying the Benjamini--Hochberg correction to the moderated
t-test p-values across all tested genes. Ties introduced by the adjustment can
appear as horizontal bands. No horizontal significance line is drawn.

```{r panel-a-volcano}
res$neglog10AdjP <- -log10(pmax(res$adj.P.Val, .Machine$double.xmin))
res$status <- "NS"
res$status[res$adj.P.Val < fdr_thresh & res$logFC >=  fc_thresh] <- "Up"
res$status[res$adj.P.Val < fdr_thresh & res$logFC <= -fc_thresh] <- "Down"
res$status <- factor(res$status, levels = c("Down", "NS", "Up"))
res$is_sig <- with(res, adj.P.Val < fdr_thresh & abs(logFC) >= fc_thresh)

up_n   <- sum(res$status == "Up", na.rm = TRUE)
down_n <- sum(res$status == "Down", na.rm = TRUE)

lab_df <- res[order(res$adj.P.Val), ][seq_len(min(20, nrow(res))), ]
x_min <- min(res$logFC, na.rm = TRUE); x_max <- max(res$logFC, na.rm = TRUE)
y_max <- max(res$neglog10AdjP, na.rm = TRUE); x_span <- x_max - x_min

fc_axis_lab <- if (transform_applied) {
  paste0("log2 Fold Change (", params$alt_label, " / ", params$ref_label, ")")
} else {
  paste0("Mean difference, input scale (", params$alt_label, " - ", params$ref_label, ")")
}

volcano <- ggplot(res, aes(x = logFC, y = neglog10AdjP)) +
  geom_vline(xintercept = -fc_thresh, color = "steelblue3", linetype = "dashed") +
  geom_vline(xintercept =  fc_thresh, color = "red3",       linetype = "dashed") +
  geom_point(aes(color = status, shape = is_sig), alpha = 0.85, size = 2) +
  scale_color_manual(values = c(Down = "steelblue3", NS = "grey70", Up = "red3")) +
  scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 17),
                     labels = c("Not sig", "Significant")) +
  # ylim keeps repelled labels out of the band the two count annotations occupy;
  # without it they collide as soon as a top gene sits near the y maximum.
  geom_text_repel(data = lab_df, aes(label = Gene), size = 3, max.overlaps = 200,
                  min.segment.length = 0, box.padding = 0.3, point.padding = 0.2,
                  ylim = c(NA, y_max * 0.88)) +
  annotate("text", x = x_min + 0.02 * x_span, y = y_max * 0.98,
           label = paste0("Up (FC >= ", fc_thresh, ", FDR < ", fdr_thresh, "): ", up_n),
           hjust = 0, vjust = 1, color = "red3", fontface = "bold") +
  annotate("text", x = x_min + 0.02 * x_span, y = y_max * 0.92,
           label = paste0("Down (FC <= -", fc_thresh, ", FDR < ", fdr_thresh, "): ", down_n),
           hjust = 0, vjust = 1, color = "steelblue3", fontface = "bold") +
  labs(title = paste0("Volcano plot: ", params$alt_label, " vs ", params$ref_label),
       x = fc_axis_lab, y = expression(-log[10]("BH-adjusted P"))) +
  theme_bw() +
  theme(plot.title = element_text(face = "bold"), legend.position = "top",
        aspect.ratio = 1) +
  coord_cartesian(clip = "off")

ggsave(paste0(out_stem, "_volcano.png"), volcano, width = 5.5, height = 5.5, dpi = 300,
       device = ragg::agg_png)
ggsave(paste0(out_stem, "_volcano.pdf"), volcano, width = 5.5, height = 5.5)
volcano
```

## 6. Panel B -- gene correlation

Pairwise Pearson correlations are calculated across cells for the genes with
the smallest BH-adjusted differential-expression p-values. The variables are
placed in PCA-based angular order. Both triangles show the same correlations:
the shaded squares encode signed `r`, while each circle's filled fraction
encodes `abs(r)`. Diagonal labels are gene symbols.

```{r panel-b-corrgram}
top_genes_cg <- head(res$Gene[order(res$adj.P.Val)], n = min(params$n_corrgram, nrow(res)))
df_cells <- as.data.frame(t(expr[top_genes_cg, , drop = FALSE]))

# Reverse corrgram's default palette so negative correlations are blue and
# positive correlations are red; white remains the midpoint at zero.
corr_palette <- colorRampPalette(c("#2166AC", "#92C5DE", "#F7F7F7",
                                   "#F4A582", "#B2182B"))

draw_corrgram_key <- function() {
  op <- par(no.readonly = TRUE)
  on.exit(par(op), add = TRUE)

  # This strip sits wholly inside the enlarged bottom outer margin. Keeping a
  # white gutter between it and the matrix prevents the key from competing with
  # the lowest row of cells or the diagonal gene labels.
  par(fig = c(0.08, 0.92, 0.005, 0.115), mar = rep(0, 4),
      new = TRUE, xpd = NA, pty = "m")
  plot.new()
  plot.window(xlim = c(-1, 1), ylim = c(0, 1), xaxs = "i", yaxs = "i")

  key_breaks <- seq(-1, 1, length.out = 201)
  rect(key_breaks[-length(key_breaks)], 0.44,
       key_breaks[-1], 0.70,
       col = corr_palette(length(key_breaks) - 1), border = NA)
  rect(-1, 0.44, 1, 0.70, border = "grey35", lwd = 0.5)

  key_ticks <- c(-1, -0.5, 0, 0.5, 1)
  segments(key_ticks, 0.39, key_ticks, 0.44, col = "grey25", lwd = 0.5)
  text(key_ticks, 0.30, labels = c("-1", "-0.5", "0", "0.5", "1"),
       cex = 0.60)
  text(0, 0.84, "Pearson correlation (r)", font = 2, cex = 0.68)
  text(0, 0.12,
       "Square: color/intensity = r   |   Circle: fill = |r|   |   Diagonal: gene symbol",
       cex = 0.62)
}

# Wrapped in a function so section 9 can re-draw it into a captured device.
# par(pty = "s") keeps the correlation grid square; on.exit restores it even if
# corrgram errors partway through.
draw_corrgram <- function() {
  op <- par(no.readonly = TRUE)
  on.exit(par(op), add = TRUE)
  par(pty = "s")
  corrgram(df_cells, order = TRUE,
           lower.panel = panel.shade, upper.panel = panel.pie, text.panel = panel.txt,
           col.regions = corr_palette,
           cor.method = "pearson", label.srt = -45, cex.labels = 0.65,
           oma = c(7.5, 4, 4.5, 4), cex.main = 1,
           main = "Gene-expression correlations")
  draw_corrgram_key()
}

ragg::agg_png(paste0(out_stem, "_corrgram_genes.png"), width = 1000, height = 1000, res = 150)
draw_corrgram()
dev.off()

pdf(paste0(out_stem, "_corrgram_genes.pdf"), width = 6.7, height = 6.7)
draw_corrgram()
dev.off()

draw_corrgram()
```

```{r panel-source-gene-csvs, include=FALSE}
# Artifact-only export: read existing panel inputs, create new tables, and write
# CSV files without modifying any plot object or source data.
local({
gene_full_names <- function(genes) {
  if (!requireNamespace("AnnotationDbi", quietly = TRUE) ||
      !requireNamespace("org.Hs.eg.db", quietly = TRUE)) {
    return(rep(NA_character_, length(genes)))
  }
  mapped <- suppressMessages(AnnotationDbi::mapIds(
    org.Hs.eg.db::org.Hs.eg.db, keys = unique(genes), keytype = "SYMBOL",
    column = "GENENAME", multiVals = "first"
  ))
  unname(mapped[genes])
}

gene_source_rows <- function(genes) {
  rows <- as.data.table(copy(res[match(genes, res$Gene), , drop = FALSE]))
  data.table(
    gene_symbol = genes,
    gene_full_name = gene_full_names(genes),
    logFC = rows$logFC,
    AveExpr = rows$AveExpr,
    P.Value = rows$P.Value,
    adj.P.Val = rows$adj.P.Val,
    status = as.character(rows$status),
    is_significant = rows$is_sig
  )
}

panel_a_genes <- gene_source_rows(lab_df$Gene)
panel_a_genes[, label_rank := seq_len(.N)]
setcolorder(panel_a_genes, c("label_rank", "gene_symbol", "gene_full_name"))
fwrite(panel_a_genes, paste0(out_stem, "_panel_A_labeled_genes.csv"))

corr_for_order <- cor(df_cells, use = "pairwise.complete.obs", method = "pearson")
corr_eigen <- eigen(corr_for_order)$vectors[, 1:2, drop = FALSE]
corr_angle <- ifelse(corr_eigen[, 1] > 0,
                     atan(corr_eigen[, 2] / corr_eigen[, 1]),
                     atan(corr_eigen[, 2] / corr_eigen[, 1]) + pi)
corr_display_genes <- colnames(corr_for_order)[order(corr_angle)]

panel_b_genes <- gene_source_rows(corr_display_genes)
panel_b_genes[, `:=`(
  display_order = seq_len(.N),
  selection_rank = match(gene_symbol, top_genes_cg)
)]
setcolorder(panel_b_genes,
            c("display_order", "selection_rank", "gene_symbol", "gene_full_name"))
fwrite(panel_b_genes, paste0(out_stem, "_panel_B_corrgram_genes.csv"))
})
```

## 7. Panel C -- PC violin

Only the two disease-status groups are plotted. An `All` category would contain
the same cells already present in Healthy and Diseased, so comparisons against
it would violate the independence assumption. The Wilcoxon annotation below is
explicitly cell-level and descriptive; donor-level inference requires a
donor-aware analysis.

```{r panel-c-violin}
rt <- data.frame(
  id         = pcoadata$sample,
  Type       = factor(pcoadata$Subtype,
                      levels = c(params$ref_label, params$alt_label)),
  Expression = pcoadata[[params$which_pc]]
)

wilcox_fit <- wilcox.test(Expression ~ Type, data = rt, exact = FALSE)
wilcox_p <- wilcox_fit$p.value
wilcox_p_label <- if (is.finite(wilcox_p) && wilcox_p > 0) {
  formatC(wilcox_p, format = "e", digits = 2)
} else {
  paste0("< ", formatC(.Machine$double.xmin, format = "e", digits = 2))
}

y_range <- range(rt$Expression, na.rm = TRUE)
y_span <- diff(y_range)
if (!is.finite(y_span) || y_span == 0) y_span <- max(abs(y_range), 1)

comparison_df <- data.frame(
  group1 = params$ref_label,
  group2 = params$alt_label,
  y.position = y_range[2] + 0.08 * y_span,
  label = paste0("Cell-level Wilcoxon p = ", wilcox_p_label)
)

group_counts <- table(rt$Type)
group_axis_labels <- setNames(
  paste0(names(group_counts), "\n(n cells = ",
         format(as.integer(group_counts), big.mark = ",", trim = TRUE), ")"),
  names(group_counts)
)

violin <- ggviolin(rt, x = "Type", y = "Expression", fill = "Type",
                   xlab = "Disease status", ylab = paste0(params$which_pc, " score"),
                   trim = TRUE, add = "boxplot",
                   add.params = list(fill = "white", width = 0.15, color = "black")) +
  stat_pvalue_manual(comparison_df, label = "label", tip.length = 0.01,
                     bracket.size = 0.5, size = 3.5) +
  scale_fill_manual(values = pal2[levels(rt$Type)], drop = FALSE) +
  scale_x_discrete(labels = group_axis_labels) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.18))) +
  theme_bw() +
  theme(axis.text = element_text(color = "black", face = "bold"),
        legend.position = "none")

out_violin <- paste0(out_stem, "_violin_", params$which_pc, "_2groups.pdf")
ggsave(out_violin, violin, width = 6, height = 4)
message("Saved: ", out_violin)
violin
```

## 8. Panel D -- PCA composite

```{r panel-d-pca}
p_main <- ggplot(pcoadata, aes(PC1, PC2)) +
  geom_point(aes(colour = Subtype, fill = Subtype), size = 4) +
  scale_color_manual(values = pal2) +
  labs(x = paste0("(PC1: ", round(pve[1] * 100, 2), "%)"),
       y = paste0("(PC2: ", round(pve[2] * 100, 2), "%)")) +
  geom_vline(xintercept = 0, linetype = "dotted") +
  geom_hline(yintercept = 0, linetype = "dotted") +
  theme_bw() +
  theme(
    panel.background = element_rect(fill = "white", colour = "black"),
    axis.title.x = element_text(colour = "black", size = 12, margin = margin(t = 5), face = "bold"),
    axis.title.y = element_text(colour = "black", size = 12, margin = margin(r = 5), face = "bold"),
    axis.text = element_text(color = "black", face = "bold"),
    plot.title = element_blank(),
    legend.title = element_blank(),
    legend.key = element_blank(),
    legend.text = element_text(color = "black", size = 9, face = "bold"),
    legend.spacing.x = unit(0.1, "cm"),
    legend.key.width = unit(0.5, "cm"),
    legend.key.height = unit(0.5, "cm"),
    legend.background = element_blank(),
    legend.box.background = element_rect(colour = "black"),
    legend.position = "inside",
    legend.position.inside = c(0.001, 0.999),
    legend.justification = c(0.0001, 1),
    aspect.ratio = 1
  )

cld_table <- function(pc) {
  f   <- aov(as.formula(paste(pc, "~ Subtype")), data = pcoadata)
  cld <- multcompLetters4(f, TukeyHSD(f))
  dt  <- aggregate(as.formula(paste(pc, "~ Subtype")), data = pcoadata, FUN = max)
  names(dt)[2] <- "value_max"
  letters_df <- as.data.frame.list(cld$Subtype)
  letters_df$Subtype <- rownames(letters_df)
  merge(letters_df, dt, by = "Subtype", all.x = TRUE)
}
text1 <- cld_table("PC1")
text2 <- cld_table("PC2")

box_theme <- theme_bw() +
  theme(panel.background = element_rect(fill = "white", colour = "black"),
        axis.text = element_blank(), axis.ticks = element_blank(),
        plot.title = element_blank(), legend.position = "none")

# after_stat() replaces the retired ..lower.. / ..ymin.. syntax; linewidth
# replaces size for line-based geoms (both deprecated since ggplot2 3.4).
p_top <- ggplot(pcoadata, aes(Subtype, PC1, fill = Subtype)) +
  geom_boxplot(outlier.shape = NA, width = 0.5, color = "black", linetype = "dotted") +
  stat_boxplot(aes(ymin = after_stat(lower), ymax = after_stat(upper)),
               outlier.shape = NA, width = 0.5) +
  stat_boxplot(geom = "errorbar", aes(ymin = after_stat(ymax)), width = 0.2, linewidth = 0.35) +
  stat_boxplot(geom = "errorbar", aes(ymax = after_stat(ymin)), width = 0.2, linewidth = 0.35) +
  geom_text(data = text1, aes(label = Letters, y = value_max),
            angle = -90, color = "black", size = 4, vjust = -0.4) +
  scale_fill_manual(values = pal2) + labs(x = NULL, y = NULL) +
  box_theme + coord_flip()

p_right <- ggplot(pcoadata, aes(Subtype, PC2, fill = Subtype)) +
  geom_boxplot(outlier.shape = NA, width = 0.5, color = "black", linetype = "dotted") +
  stat_boxplot(aes(ymin = after_stat(lower), ymax = after_stat(upper)),
               outlier.shape = NA, width = 0.5) +
  stat_boxplot(geom = "errorbar", aes(ymin = after_stat(ymax)), width = 0.2, linewidth = 0.35) +
  stat_boxplot(geom = "errorbar", aes(ymax = after_stat(ymin)), width = 0.2, linewidth = 0.35) +
  geom_text(data = text2, aes(label = Letters, y = value_max),
            angle = 0, color = "black", size = 4, vjust = -0.4) +
  scale_fill_manual(values = pal2) + labs(x = NULL, y = NULL) +
  box_theme

p_stat <- ggplot() +
  geom_text(aes(x = 0, y = 0.1,
                label = paste("PERMANOVA",
                              paste0("df = ", otu.adonis$Df[1]),
                              paste0("R2 = ", round(otu.adonis$R2[1], 3)),
                              perm_p_label,
                              paste0("perms = ", params$permutations),
                              sep = "\n")),
            size = 3.5, color = "black", fontface = "bold") +
  theme_bw() +
  theme(panel.background = element_rect(fill = "white", colour = "black"),
        axis.title = element_blank(), axis.ticks = element_blank(),
        axis.text = element_blank(), plot.title = element_blank(),
        legend.position = "none")

final_plot <- p_top + p_stat + p_main + p_right +
  plot_layout(heights = c(1, 4), widths = c(4, 1), ncol = 2, nrow = 2)

ggsave(paste0(out_stem, "_PCA_panel.png"), final_plot, width = 6, height = 6, dpi = 300,
       device = ragg::agg_png)
ggsave(paste0(out_stem, "_PCA_panel.pdf"), final_plot, width = 6, height = 6)
final_plot
```

## 9. Assembled A-B-C-D figure

The four panels above are also written as separate files, because that is what
you want when one panel needs re-running. This chunk additionally emits the
single-row composite with A/B/C/D tags.

Two things make this less trivial than `pA | pB | pC | pD`:

* Panel B is **base graphics** (`corrgram` draws to the device and returns
  nothing composable), so patchwork cannot take it as an operand. It must be
  captured, which is why it is drawn through the `draw_corrgram()` closure
  rather than inline. `wrap_elements(full = ~ draw_corrgram())` is the obvious
  route and it **silently produces a near-empty panel** -- corrgram lays its
  grid out with base `par(mfrow)`, and that state does not survive the
  gridGraphics replay. `ggplotify::as.ggplot(function() ...)` captures it
  correctly. Verified by output size: the `wrap_elements` route wrote 17 KB,
  the `ggplotify` route 95 KB for the same figure.
* Panel D is itself a patchwork of four sub-plots. Passing it directly would let
  `tag_levels = "A"` recurse into its children and tag them A-G. Wrapping it
  makes it opaque to tagging, so it stays a single "D".

Everything is re-rendered into a quarter-width cell here, and two things do not
survive that cleanly:

* fixed-size text shrinks relative to its box -- the PERMANOVA block in panel D
  clips its header and permutation count;
* panel B's captured region stretches to the cell, so the `panel.pie` circles
  become ellipses. `par(pty = "s")` does not prevent this, because the capture
  device is created at the cell's aspect, not a square one.

Neither affects the standalone files, which are correct. For a publication
figure, assemble from the per-panel PDFs, where each panel keeps its own scale;
this composite is for quick review of all four at once. Set `composite: false`
to skip it -- it roughly triples total runtime, since the volcano's ~29k points
are rasterised and vectorised a second time at 22 inches wide.

```{r figure-abcd, eval = params$composite}
panel_b <- ggplotify::as.ggplot(function() draw_corrgram())
panel_d <- wrap_elements(final_plot)

figure_abcd <- (volcano | panel_b | violin | panel_d) +
  plot_annotation(tag_levels = "A") &
  theme(plot.tag = element_text(face = "bold", size = 18))

ggsave(paste0(out_stem, "_figure_ABCD.png"), figure_abcd,
       width = 22, height = 5.5, dpi = 300, limitsize = FALSE,
       device = ragg::agg_png)
ggsave(paste0(out_stem, "_figure_ABCD.pdf"), figure_abcd,
       width = 22, height = 5.5, limitsize = FALSE)
message("Composite written to: ", paste0(out_stem, "_figure_ABCD.{png,pdf}"))
```

## 10. Optional -- donor-level pseudobulk

Off by default (`params$do_pseudobulk`). Everything above tests cells as if they
were replicates. This chunk rebuilds the contrast with donors as the unit.

Two hard constraints, both from the data rather than from this code:

* The `foranalysis_*` CSVs carry no donor, batch, or dataset column. Provenance
  lives only in `labels.csv`, joined by **positional convention**. That
  convention breaks whenever `select_contrast()` in the Python workflow
  populates `excluded`, because those cells appear in `labels.csv` and in
  neither matrix. The guard below refuses to proceed rather than silently
  misalign.
* True pseudobulk sums **raw counts** per donor and normalises afterwards. Only
  CP10K is exported, so this averages CP10K instead. That is a common
  approximation but it is not count-based pseudobulk, and it cannot recover the
  depth weighting. Fixing it properly requires exporting counts on the Python
  side.

```{r pseudobulk, eval = params$do_pseudobulk}
lab <- fread(file.path(params$session_dir, params$labels_csv))

n_alt <- ncol(mat_alt); n_ref <- ncol(mat_ref)
if (nrow(lab) != n_alt + n_ref) {
  stop(sprintf(paste0(
    "labels.csv has %d rows but the two matrices hold %d cells. The positional ",
    "join is invalid -- most likely select_contrast() excluded one or more ",
    "classes. Re-export with the excluded cells, or add an explicit cell id."),
    nrow(lab), n_alt + n_ref))
}

# Verify the assumed [alt block; ref block] ordering instead of trusting it.
frac_norm <- function(v) mean(grepl("^(normal|healthy)$", tolower(trimws(v))))
f_first <- frac_norm(head(lab$disease, n_alt))
f_last  <- frac_norm(tail(lab$disease, n_ref))
if (!(f_first < 0.05 && f_last > 0.95)) {
  stop(sprintf(paste0(
    "labels.csv row order does not match [%s block; %s block] ",
    "(normal fraction: first block %.2f, second block %.2f). Refusing to join."),
    params$alt_label, params$ref_label, f_first, f_last))
}

meta <- data.frame(
  cell  = colnames(expr),
  group = group,
  donor = c(tail(lab$donor_id, n_ref), head(lab$donor_id, n_alt)),
  stringsAsFactors = FALSE
)
# expr columns are [ref block; alt block]; lab rows are [alt block; ref block].

keep <- !is.na(meta$donor) & tolower(trimws(meta$donor)) != "unknown"
message(sprintf("Cells with a usable donor_id: %d of %d (%.0f%% dropped)",
                sum(keep), nrow(meta), 100 * mean(!keep)))

meta_k <- meta[keep, ]
expr_k <- expr[, keep, drop = FALSE]

donor_key <- paste(meta_k$group, meta_k$donor, sep = "|")
pb <- sapply(split(seq_len(ncol(expr_k)), donor_key),
             function(idx) rowMeans(expr_k[, idx, drop = FALSE]))

pb_group <- factor(sub("\\|.*$", "", colnames(pb)),
                   levels = c(params$ref_label, params$alt_label))
message(sprintf("Pseudobulk units: %d (%s n=%d, %s n=%d)",
                ncol(pb),
                params$ref_label, sum(pb_group == params$ref_label),
                params$alt_label, sum(pb_group == params$alt_label)))

pb_design <- model.matrix(~ 0 + pb_group)
colnames(pb_design) <- make.names(levels(pb_group))
pb_cm  <- makeContrasts(contrasts = contrast_str, levels = pb_design)
pb_fit <- eBayes(contrasts.fit(lmFit(pb, pb_design), pb_cm), robust = TRUE, trend = TRUE)
pb_res <- topTable(pb_fit, coef = 1, number = Inf, sort.by = "P")
pb_res$Gene <- rownames(pb_res)
pb_res <- pb_res[, c("Gene", "logFC", "AveExpr", "t", "P.Value", "adj.P.Val", "B")]

fwrite(pb_res, paste0(out_stem, "_table_pseudobulk_donor.csv"))
message("Donor-level FDR < ", fdr_thresh, ": ",
        sum(pb_res$adj.P.Val < fdr_thresh, na.rm = TRUE),
        "  (cell-level was ", sig_count, ")")
```

## 11. Optional -- heatmap (disabled)

Not part of panels A-D. Kept for reference and disabled: `pheatmap` with
`clustering_method = "complete"` on this many columns builds an
n-cells x n-cells distance matrix, is slow, produces an illegible image at that
width, and -- because a knitr error aborts the document by default -- would take
panels A, B, C and D down with it.

```{r heatmap, eval = FALSE}
top_genes_hm <- head(res$Gene[order(res$adj.P.Val)], n = min(20, nrow(res)))
ann <- data.frame(Group = group)
rownames(ann) <- colnames(expr)

ragg::agg_png(paste0(out_stem, "_heatmap.png"), width = 900, height = 1100, res = 130)
pheatmap(expr[top_genes_hm, , drop = FALSE],
         scale = "row", clustering_method = "complete",
         annotation_col = ann, show_rownames = TRUE, show_colnames = FALSE,
         main = paste0("Top ", length(top_genes_hm), " DE genes"))
dev.off()
```

## Statistical scope

**What the code above is now internally consistent about:** the fold-change
scale, the significance threshold drawn on panel A, the contrast direction, the
execution order, and the permutation p-value reporting.

**What it does not fix.** Every column is a single cell. limma, the Wilcoxon
comparisons in panel C, and the PERMANOVA in panel D all treat cells as
independent replicates. They are not: cells from one donor are correlated, and
the effective sample size is the number of donors, not the number of cells. This
inflates significance by orders of magnitude and is why panel C reports
`p = 1.93e-187`. For any inferential claim, use section 9 or a mixed-effects
model with a donor random effect. The cell-level results are a ranking and a
visualisation, not evidence of a population-level difference.

**Rendering.** `pandoc` is not on this machine, so `rmarkdown::render()` runs
every chunk and then fails at the HTML conversion step. Two options.

With pandoc (`conda install -c conda-forge pandoc`):

```
module load r/4.5.2
Rscript -e 'rmarkdown::render("casestudy_panels.Rmd", params = list(
  session_dir = "/path/to/webapp/sessions/<id>",
  ref_csv     = "foranalysis_combined_normal_df_<Disease>.csv",
  alt_csv     = "foranalysis_combined_disease_df_<Disease>.csv"))'
```

Without pandoc, use `knitr::knit()` -- it honours `eval=` and produces a `.md`
plus all the figure files. `params` must be defined first, because `knit()` does
not build it from the YAML the way `render()` does:

```
module load r/4.5.2
Rscript -e 'params <- list(session_dir = "/path/to/webapp/sessions/<id>",
                           ref_csv = "...", alt_csv = "...", ref_label = "Healthy",
                           alt_label = "Diseased", out_dir = NULL,
                           out_prefix = "DE_results", input_scale = "linear_cp10k",
                           fdr_thresh = 0.05, fc_thresh = 0.25, permutations = 999,
                           n_corrgram = 20, n_pca = 100, which_pc = "PC1",
                           do_pseudobulk = FALSE, labels_csv = "labels.csv")' \
        -e 'knitr::knit("casestudy_panels.Rmd", output = "casestudy_panels.md")'
```

Do **not** use `knitr::purl()` here: it ignores `eval=FALSE` (so it would run the
disabled heatmap chunk) and it errors on `eval = params$do_pseudobulk` because
`params` does not exist at purl time.
